---
title: "Final Exam"
author: "Dawid Dąbkowski"
date: "2 February 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PRE-PROCESSING

We start by setting the random seed and loading useful libraries.

```{r, message=F, warning=F}
set.seed(02022018)
require(RCurl)
require(gdata)
library(dplyr)
library(magrittr)
library(ggplot2)
library(MASS)
#library(MVN) #multivariate normality test
#library(MASS)
#library(caret)
```

Now we load our data frames for the exercises.

```{r}
urls <- c("https://www.mimuw.edu.pl/~noble/courses/SDA/data/ushighways.txt",
         "https://www.mimuw.edu.pl/~noble/courses/SDA/data/PET.txt",
         "https://www.mimuw.edu.pl/~noble/courses/SDA/data/ozone.csv",
         "https://www.mimuw.edu.pl/~noble/courses/SDA/data/pendigits.txt",
         "https://www.mimuw.edu.pl/~noble/courses/SDA/data/carmarks.txt",
         "https://www.mimuw.edu.pl/~noble/courses/SDA/data/primate.scapulae.txt")
ushighways <- read.table(urls[1], header=T)
#download.file("https://www.mimuw.edu.pl/~noble/courses/SDA/data/bodyfat2.XLS", destfile="bodyfat2.xls")
bodyfat <- read.xls("bodyfat2.xlsx")
#data(yarn, package="pls")
yarn <- read.table(urls[2], header=T)
ozone <- read.csv(urls[3])
pendigits <- read.table(urls[4])
carmarks <- read.table(urls[5], sep=";", header=T)
scapular <- read.table(urls[6], sep=" ", header=T)
```

## EXERCISE 1

Let us take a look at data.

```{r}
head(ushighways)
```

We first draw some histograms with different window widths.

```{r}
ggplot(ushighways, aes(x=Approx.Miles)) + 
    geom_histogram(size=2, alpha=0.05, binwidth=20, boundary=0, 
                   aes(color="20", y=..density..)) + 
    geom_histogram(size=1.5, alpha=0.05, binwidth=10, boundary=0, 
                   aes(color="10", y=..density..)) +
    geom_histogram(size=1, alpha=0.05, binwidth=5, boundary=0, 
                   aes(color="5", y=..density..)) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5)) + ggtitle("Histograms") +
    guides(color=guide_legend(title="Window width")) + 
    theme(legend.position=c(0.9,0.8), legend.background=element_rect(fill=alpha('blue',0)))
```

The second picture with window width of 10 miles looks quite good. It is well-splitted and shows the  monotonicity of the distribution.

Now we will use UCV, BCV and SJPI estimators for window width to plot densities with gaussian kernel.

```{r}
sjpi <- width.SJ(ushighways$Approx.Miles, method="dpi")
ggplot(ushighways, aes(x=Approx.Miles)) + 
    geom_density(fill="grey", size=0.8, alpha=0.05, bw="ucv", aes(color="UCV")) + 
    geom_density(fill="grey", size=0.8, alpha=0.05, bw="bcv", aes(color="BCV")) +
    geom_density(fill="grey", size=0.8, alpha=0.05, bw="sj", aes(color="SJ")) +
    geom_density(fill="grey", size=0.8, alpha=0.05, bw=sjpi, aes(color="SJPI")) +
    theme_bw() + theme(plot.title = element_text(hjust = 0.5)) + 
    ggtitle("Density estimators") + guides(color=guide_legend(title="Window width")) + 
    theme(legend.position=c(0.9,0.8), legend.background=element_rect(fill=alpha('blue',0)))
```

The BCV estimator (red one) looks quite well-fitted and monotonic. The other kernel choices (triangular, cosine etc.) gave very similar results.

## EXERCISE 2

Let us take a look at data.

```{r}
head(bodyfat)
```

## EXERCISE 3

Let us take a look at data.

```{r}
head(yarn)
```

## EXERCISE 4

Let us take a look at data.

```{r}
head(ozone)
```

## EXERCISE 5

Let us take a look at data.

```{r}
head(pendigits)
```

## EXERCISE 6

Let us take a look at data.

```{r}
head(carmarks)
```

## EXERCISE 7

Let us take a look at data.

```{r}
head(scapular)
```

## EXERCISE 8

Let us take a look at data.

```{r}
head(scapular)
```